---
title: "Class 8: Mini-Project"
author: "Mai Tamura (PID: 18594079)"
format: pdf
toc: true
---

## Background

The goal of this mini-project is for you to explore a complete analysis using the unsupervised learning techniques covered in class. You’ll extend what you’ve learned by combining PCA as a preprocessing step to clustering using data that consist of measurements of cell nuclei of human breast masses. This expands on our RNA-Seq analysis from last day.

The data itself comes from the Wisconsin Breast Cancer Diagnostic Data Set first reported by K. P. Benne and O. L. Mangasarian: “Robust Linear Programming Discrimination of Two Linearly Inseparable Sets”.

Values in this data set describe characteristics of the cell nuclei present in digitized images of a fine needle aspiration (FNA) of a breast mass.


## Data Import

Data was downloaded from the class website as a CSV file.

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names=1)
head(wisc.df)
```

The first column `diagnosis` in the expert opinion on the sample (i.e. patient FNA).

```{r}
wisc.df$diagnosis
```

Remove the diagnosis column from data sebsequent analysis.

```{r}
wisc.data <- wisc.df[,-1]
dim(wisc.data)
```

Store the diagnosis as a vector for use later when we compare our result to those from experts in the field.

```{r}
diagnosis <- wisc.df$diagnosis
```

> Q1. How many observations are in this dataset?

There are `r nrow(wisc.data)` observations/patients in the dataset.

> Q2. How many of the observations have a malignant diagnosis?

```{r}
table(diagnosis)
```

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
#colnames(wisc.data)
length( grep("_mean", colnames(wisc.data)) )
```


## Principal Component Analysis (PCA)

The `prcomp()` function to do PCA has a `scale=FALSE` default. In general we nearly always want to set this to TRUE so our analysis is not dominated by columns/variables in our dataset that have high standard deviation and mean when compared to others just because the units of measurements are on different units/scales.

```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE)
summary <- summary(wisc.pr)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

```{r}
summary$importance["Proportion of Variance","PC1"]
```

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

```{r}
which( summary$importance["Cumulative Proportion", ] >= 0.70)[1]
```

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

```{r}
which( summary$importance["Cumulative Proportion", ] >= 0.90)[1]
```

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

```{r}
biplot(wisc.pr)
```

The main PC result figure is called a "score plot" or "PC plot" or "ordination plot"...

```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```


## PCA Scree-plot

```{r}
pr.var <- wisc.pr$sdev^2
#head(pr.var)

pve <- pr.var / sum(pr.var)

plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```


### Communicating PCA results

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation["concave.points_mean", "PC1"]
```

> Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

```{r}
summary <- summary(wisc.pr)
which(summary$importance["Cumulative Proportion", ] >= 0.80)[1]
```

## Hierarchical clustering

Just clustering the original data is not very informative or helpful.

```{r}
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist)
```

> Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h=19, col="red", lty=2)
```

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
table(wisc.hclust.clusters)
```

```{r}
table(wisc.hclust.clusters, diagnosis)
```

## Combining methods (PCA and Clustering)

Clustering the original data was not very productive. The PCA results looked promising. Here we combine these methods by clustering from our PCA results. In other words, "clustering in PC space"...

```{r}
## take the first 3 PCs
dist.pc <- dist( wisc.pr$x[,1:3] )
wisc.pr.hclust <- hclust(dist.pc, method="ward.D2")
```

```{r}
plot(wisc.pr.hclust)
abline(h=70, col="red")
```

To get our clustering membership vector (i.e. our main clustering result) we "cut" the tree at a desired height or to yield a desired number of "k" group.

```{r}
grps <- cutree(wisc.pr.hclust, h=70)
table(grps)
```

How does this clustering grps compare to the expert?

```{r}
table(grps, diagnosis)
```

Sensitivity: TP/(TP+FN)
Specificity: TN/(TN+FN)

## Prediction

We can use our PCA model for prediction with new input patient samples.

```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```


