---
title: "Class 10: Halloween Mini-Project"
author: "Mai Tamura (PID: A18594079)"
format: pdf
toc: true
---

## Bachground

As it is nearly Halloween and the half way point in the quarter, let's do a mini project to help us figure out the best candy!

Our come from the 538 website and is available as a CSV file:

## Data Import

```{r}
candy <- read.csv("candy-data.csv", row.names = 1)
head(candy)
```

```{r}
flextable::flextable(head(candy, 10))
```


> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```
```{r}
candy |>
  nrow()
```
```{r}
library(tidyverse)

candy %>%
  nrow
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Haribo Gold Bears", ]$winpercent
```

```{r}
library(dplyr)

candy |>
  filter(rownames(candy) == "Haribo Gold Bears") |>
  select(winpercent)
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

## Quick Overview of the Dataset

```{r}
skimr::skim(candy)
```


> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The winpercent is on a 0-100 scale, while the rest are 0-1 scale.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

Zero represents that the candy does not contain chocolate, while one represents that it does.

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins = 20)
```

> Q9. Is the distribution of winpercent values symmetrical?

```{r}
ggplot(candy) +
  aes(winpercent) +
  geom_density()
```

> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```

```{r}
summary(candy$winpercent)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# 1. Find all chocolate candy in the dataset
# 2. Find their winpercent values
# 3. Calculate the mean of these values
# 4-6. Do the same for fruity candy
# 7. Compare mean winpercents of chocolate vs fruity
# 8. Pick the highest as the winner

choc.inds <- candy$chocolate == 1
choc.win <- candy[choc.inds, ]$winpercent
choc.mean <- mean(choc.win)
choc.mean

fru.inds <- candy$fruity == 1
fru.win <- candy[fru.inds, ]$winpercent
fru.mean <- mean(fru.win)
fru.mean
```

```{r}
mean(candy[candy$chocolate==1, ]$winpercent)
mean(candy[candy$fruity==1, ]$winpercent)
```

```{r}
candy |>
  filter(chocolate==1) |>
  select(winpercent)
```

> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fru.win)
```

## Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

```{r}
candy |>
  arrange(winpercent) |>
  head(5)
```

```{r}
x <- c(5, 1, 10, 4)
#sort(x)
order(x)
#(candy$winpercent)
```

```{r}
ord.inds <- order(candy$winpercent)
head(candy[ord.inds,], 5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy |>
  arrange(winpercent) |>
  tail(5)
```

```{r}
candy |>
  arrange(-winpercent) |>
  head(5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(x=winpercent, 
      y=reorder(rownames(candy), winpercent)) +
  geom_col()
```

Add some color based on the type of candy

```{r}
my_cols <- rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] <- "chocolate"
my_cols[as.logical(candy$bar)] <- "brown"
my_cols[as.logical(candy$fruity)] <- "pink"
my_cols
```

```{r}
ggplot(candy) +
  aes(x=winpercent, 
      y=reorder(rownames(candy), winpercent)) +
  geom_col(fill=my_cols)
```

> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets.

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starburst.

## Winpercent and Pricepercent

A plot with both variables/columns winpercent and pricepercent

```{r}
my_cols[as.logical(candy$fruity)] <- "red"

library(ggrepel)

ggplot(candy) +
  aes(x=winpercent, 
      y=pricepercent,
      label=rownames(candy)) +
  geom_point(col=my_cols) +
  geom_text_repel(col=my_cols, max.overlaps = 7)
```

## Exploring the Correlation Structure

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

## Principal Component Analysis

The function to use is called `prcomp()` with an optional `scale=T/F` argument.

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

Our main PCA result figure

```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point(col=my_cols) +
  geom_text_repel(col=my_cols)
```

We should also examine the variable "loadings" or contributions of the original variables to the new PCs

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```

```{r}
p <- ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point(col=my_cols) +
  geom_text_repel(col=my_cols)
```

Interactive plots that can be zoomed on and "brushed" over can be made with the **plotly** package.

```{r}
library(plotly)

#ggplotly(p)
```


